How Consistent are LLMs in Applying Handcrafted Language Features

### 2024-06-01 article/main.tex
fill out hlfs, datasets and correct existing text

\textbf{Large language models (LLMs)} have are now a commonplace research topic.
Because language models such as BERT~\cite{devlin2018bert} or GPT~\cite{gpt-2018,gpt2-2019,
gpt3-2020} constantly advance the state of the art for many natural language
processing tasks, it has become natural to want to evaluate these models
performance on ever more tasks.
We know from multiple surveys~\cite{minaee2024llmsurvey,zhao2023survey} and
benchmarks~\cite{papcode2024hellaswag,chiang2024chatbot} that large language
models (especially the more popular ones) are very good at following
instructions.
We can also intuit that training LLMs on diverse data (for instance the
Pile~\cite{gao2020pile}) uniquely qualifies them to produce text in a wide
variety of styles.

\textbf{Text style transfer (TST)} is defined as the ``task of transforming the
stylistic manner in which a sentence is written, while preserving the meaning of
the original sentence''~\cite{tst-review-2021}.
This definition can be extended, to entire articles or corpuses containing
multiple articles without fear of text style transfer losing its meaning.
The task of transfering text style is an old preocupation for the natural
language processing and computational linguistics communities which has picked
up interest again in the context of the advancements in the field of deep
learning and particularly since the inception of the transformer architecture.

\textbf{Handcrafted linguistic features (HLFs)} are single numerical values produced by a
uniquely identifiable method on any natural language~\cite{lftk-2023}. More
often than not, these features are easy to compute and intuitively very telling
of the style in which a text is written.

Let's take the number of exclamations in a text as an example.
This particular HLF is one of the easier computational linguistics features to
compute.
Yet high values of this particular HLF often provide a clear signal for
inflamatory news writing style, spam or an alert narrative.
Similar observations have been made about other characteristics of written text
that might be computed as HLFs~\cite{hovy1987generating}.
This gives rise to the hope that text style can be sufficiently, although not
completely, described through the use of HLFs.
What's most appealing about HLFs is their idempotent nature.
They could potentially be used both for interpreting style and generating text
with a given style.

From the above ideas, a couple of questions arise naturally.
Firstly, is it possible to influence an LLMs writing style by instructing it to
generate text in accordance to certain HLFs?
This problem of controlled text generation has implications for text style
transfer and authorship evaluation.
It might also constitute a cheaper and less intrusive approach than fine tuning
an LLM.
Fine tuning an LLM incurs some loss of generality~\cite{yang2024unveiling}.

Secondly, we ask if it's possible to tell just how well the most prominent LLMs
of today are able to generate text according to instructions concerning HLFs?
The answer to this question could hint at a relatively simple and accurate way
of at least partially evaluating LLM performance on the text style transfer
task.

@@

From a linguistic perspective, text style can be thought of as the form used for
delivering meaning~\cite{tst_sigkdd_review_2022}.
The ample literature on the subject seems to agree that style depends on context
and author choice vis-a-vis a communication
goal~\cite{mcdonald1985computational,hovy1987generating}.
The author's choice is available at all levels: lexical, syntactic and
semantic~\cite{dimarco1994model}.
These findings provide the hope and evidence that the current paper is useful
especially when it comes to the influence of syntactic structure over text
style~\cite{chomsky2002syntactic}.

The idea of constructing style from fine-grained aspects is not new and there
are excellent studies on the subject. The StylePTB authors (StylePTB is one of
the benchmark datasets used in this paper) explain how this dataset leverages
lexical, syntactic, semantic and thematic aspects to allow verifying text style
transfer more thoroughly than previous methods~\cite{lyu-etal-2021-styleptb}.
In the current paper we use this dataset merely to instruct LLMs to rewrite text
from it, without actually measuring style transfer quality.

The body of work on HLFs is extensive.
It is well referenced and synthesised by Lee and Lee~\cite{lftk-2023}.
To our knowledge there is no work that connects HLFs with LLMs in the manner
described in this paper.
Most other literature focuses on achieving results on connected tasks, such as
assessing text readability~\cite{lee-etal-2021-pushing}.
This paper differs from other tasks that aim to control the style of generated
text in at least two significant ways.
Firstly, this paper is simply about feeding well crafted user prompts to an LLM
instead of diving into deeper concepts such as the transformer architecture,
deep learning or even neural networks.
Secondly, it focuses on manipulating the style attributes of the text generated
by an LLM squarely using HLFs and measuring the change in output.
There is no ulterior motive.

Finally, this paper documents a form of controllable text generation (CTG)~\cite{zhang-ctg-2022}.
Hightened recent interest in CTG and especially in benchmarking LLMs in the
context of CTG~\cite{chen2024benchmarking} is particularly relevant for this
paper.
Such research alleviates the burden of proof regarding the level to which
LLMs respond well to diversified instructions for controlled text generation in
general.
The novelty of this paper lies both in basing the control instructions on
HLFs and in using those HLFs to evaluate the performance of the LLM.

@@

Our experiment consists of analysing the behavior of an LLM when asked to
generate new text based on a given text and and instructions derived from HLF
statistics computed on a benchmark dataset.
To perform the experiment we must know which LLMs we are going to evaluate and
what benchmark datasets we are going to use.
We must also select the HLFs for our experiment.

Before answering those questions, let us outline the experimental process.

The first task in the experiment is to establish a baseline.
We do so by instructing the target LLM to generate text retaining the meaning of
the input text without any further instruction.
We repeat the text generation step 10 times.
For each text generation, we compute the following statistics for each of our
chosen HLFs:

@@

A second baseline is obtained by following the above process with a new prompt
that builds upon the first one by adding the instruction to use example
text from our benchmark dataset of choice.
The examples are five texts chosen randomly once for each benchmark dataset.
These example texts are the same for all evaluated LLMs.

The next step is to devise HLF prompts derived from both baseline prompts.
The HLF prompts contain additional instructions to generate text that would
comply with certain HLF values.
These HLF values are obtained by computing the statistics outlined above on
the entirety of the target benchmark dataset.
We then repeat the process through which we created the baselines for each of
the HLF prompts.

To answer our first question, we compare the first baseline with the results
obtained by using the corresponding modified HLF prompt.
Our reasoning is that if the LLM were influenced by the HLF prompt, the output
would show different HLF values compared to the baseline.

To answer the question of how much LLMs are influenced by being prompted with
instructions about HLFs, we turn our attention to the second baseline.
Given that the LLM now also has example articles, it can freely immitate those
articles when generating the text.
If there is a consistent difference between the HLF and non-HLF prompt, this
difference can only stem from the HLF instructions.
We perform this second experiment with input text from both within and outside
the corpus.

@@

\begin{table}[ht]
    \setlength\tabcolsep{6pt}
    \centering
\begin{tabular}{@{}lllll@{}}\toprule
Name                      & Parameters           & Context Size        & Elo Score & License Type                                      \\ \toprule
OpenAI GPT-4o             & undisclosed          & 128000              & 1287      & Proprietary                                       \\
Google Gemini 1.5 Pro     & undisclosed          & 1048576             & 1268      & Proprietary                                       \\
Anthropic Claude 3 Opus   & undisclosed          & 200000              & 1248      & Proprietary                                       \\
Meta Llama 3-70B-Instruct & 70 Billion           & 8000                & 1208      & Llama 3 Community                                 \\
Cohere Command R+         & 104 Billion          & 128000              & 1189      & CC-BY-NC 4.0 \\
                          &                      &                     &           & with Acceptable Use\\
                          &                      &                     &           & Addendum\\ \bottomrule
\end{tabular}
\caption{Experiment LLMs}\label{table-llm}
\end{table}

@@

On the other hand, we need to look at which handcrafted linguistic features are
relevant for our experiment.
In our opinion, good handcrafted linguistic features for the task of text style
transfer are those features that can reasonably be thought to affect one of the
coordinates by which people judge the style of text.
These coordinates are well enumerated both in~\cite{tst-review-2021} and
in~\cite{tst-survey-2022}.

In order to build upon the previous work done regarding computing handcrafted
linguistic features, we will rely only on the features available in
LFTK~\cite{lftk-2023}.

Based on the above two selection criteria, we have selected the handcrafted
linguistic features in <INSERT TABLE HERE>.

@@

- Are there benchmark datasets for text style transfer? (STYLEPTB + Yelp Reviews)

@@

- link to text from within the corpus
- link to text from outside the corpus
- link to each system prompt (2 baselines, 2 HLF prompt templates)

@@

Ox - the number of the generation 1..10
Oy - HLF value
Continuous grey line - baseline metric
continuous grenat line - hlf metric
grey around grenat line - corpus target zone
dotted black lines - corpus min, max
continuous black line - corpus avg

\subsection{Baseline}

Yelp Results

STYLEPTB Results

@@

Yelp Results
StylePTB Results

>>>>

\textbf{Large language models (LLMs)} have are a commonplace research topic
today.
Because language models such as BERT~\cite{devlin2018bert} or
GPT~\cite{gpt-2018,gpt2-2019,gpt3-2020} constantly advance the state of the art
on many natural language processing tasks, it has become natural to want to
evaluate the performance of these generalist models on ever more specialized
tasks.
We know from multiple surveys~\cite{minaee2024llmsurvey,zhao2023survey} and
benchmarks~\cite{papcode2024hellaswag,chiang2024chatbot} that large language
models (especially the more popular ones) are good at following instructions.
We can also intuit that training LLMs on diverse data (for instance the
Pile~\cite{gao2020pile}) uniquely qualifies them to produce text in a wide
variety of styles.

\textbf{Text style transfer (TST)} is defined as the ``task of transforming the
stylistic manner in which a sentence is written, while preserving the meaning of
the original sentence''~\cite{tst-review-2021}.
This definition can be extended to entire articles or corpuses of text because
these are also the object of linguistic style and
stylistics~\cite{lugea2023stylistics}.
The task of transfering text style is an old preocupation for the natural
language processing and computational linguistics communities which has picked
up interest again in the context of the advancements in the field of deep
learning and particularly since the inception of the transformer architecture.

\textbf{Handcrafted linguistic features (HLFs)} are single numerical values
produced by a uniquely identifiable method on any natural
language~\cite{lftk-2023}.
More often than not, these values are easy to compute while the methods used for
obtaining them are usually idempotent.
This makes them very attractive because of the dual purpose that they can serve.
On the one hand, they can be used to instruct the generation of text (``write
no more than 10 sentences'') while on the other they can be used to analyse text
and, therefore, verify generated text.

Some HLFs statistically describe text in terms of its basic syntactical and
morphological components.
Knowing this we could follow Noam Chomsky's ideas~\cite{chomsky2002syntactic}
which imply that the style of a text is influenced by its underlying syntactic
structures.
Traditionally, linguistic style was thought to be the result of morphological
and syntactic arrangements~\cite{lugea2023stylistics}.
More recent schools of thought on linguistic style find semantics and pragmatics
to be inextricably linked to author style~\cite{verma2019lexical}, but argue
that a great deal can still be inferred from lexicon, syntax and a `surface'
level which contains basic HLFs such as the number of words in a sentence.

For example, a high number of exclamations in a text is often a revealing
signal of an inflamatory news writing style, of spam or of alert narration or
dialogue.
Similar observations have been made about other characteristics of written text
that might be computed as HLFs~\cite{hovy1987generating,lugea2023stylistics}.
Combining multiple HLFs that are revelatory of linguistic style could provide us
with an easily interpretable and verifiable stylistic profile.

These apparently disparate strands of thought lead us to the two research
questions we'd like to answer in this paper.

Firstly, \textbf{is it possible to influence an LLM's writing style by
    instructing it to generate text that complies to certain HLFs}?
This problem of controlled text generation has implications for text style
transfer and authorship verification.
Answering this question might point out cheaper, less intrusive and more user
friendly solutions than fine tuning large language models for these tasks.
Fine tuning LLMs also incurs some loss of generality~\cite{yang2024unveiling},
going directly against the purpose of having a generic language model.

Secondly, we ask \textbf{if it's possible to quantify how closely an LLM is able
    to follow prompts concerning the compliance of the generated text to HLFs?}
The answer to this question could hint at a relatively simple and accurate way
of evaluating the quality of (partial) text style transfer using a large
language model.

@@

Linguistic style, then, is the sum of identifiable language choices manifest in
a text made from the language system by the text
producer~\cite{lugea2023stylistics}.
Another way to think about it is that style is the form used for delivering
meaning~\cite{tst_sigkdd_review_2022}.
The ample literature on the subject seems to agree that style depends on context
and author choice vis-a-vis a communication
goal~\cite{mcdonald1985computational,hovy1987generating}.
The author's choice is available at all linguistic levels, including the
morphological, lexical, syntactic, semantic and pragmatic
levels~\cite{dimarco1994model,lugea2023stylistics}.
The influence of syntactic structure over text style is apparent also from a
formalist perspective~\cite{chomsky2002syntactic}.
This work grounds our choice of handcrafted linguistic features related to
linguistic style.

The idea of constructing style from fine-grained aspects is not new and there
are excellent studies on the subject.
The StylePTB authors explain how this dataset leverages lexical, syntactic,
semantic and thematic aspects to allow verifying text style transfer more
thoroughly than previous methods~\cite{lyu-etal-2021-styleptb}.
Besides providing the understanding required to distinguish between semantic
transfer and other forms of style transfer, the dataset itself is useful for
evaluating HLFs that function at the sentence level.

The body of work on HLFs is extensive.
It is well referenced and synthesised by Lee and Lee~\cite{lftk-2023}.
To our knowledge there is no work that connects HLFs with LLMs in the manner
described in this paper.
HLFs are used in the context of other, connected tasks such as assessing text
readability~\cite{lee-etal-2021-pushing}.

In terms of controlling the style of the generated text, our approach differs
from others in at least two significant ways: simplicity and focus.
We are constructing well crafted user prompts for an LLM instead of diving into
more complex artificial intelligence solutions.
Our focus is on controlling text generation using HLFs and measuring the change
relative to a baseline obtained without using HLFs in the generation process.
We are not discussing the effectiveness of style transfer, author verification
or other tasks (yet).

Finally, this paper documents a form of controllable text generation
(CTG)~\cite{zhang-ctg-2022}.
Hightened recent interest in CTG and especially in benchmarking LLMs in the
context of CTG~\cite{chen2024benchmarking} is particularly relevant for this
paper.
Such research alleviates the burden of proof regarding the level to which
LLMs respond well to diversified instructions for controlled text generation in
general.
Compared to the existing work, our prompts are guided by HLFs.
Furthermore, we use those same HLFs to evaluate the performance of the LLM.

@@

Our experiment consists of analysing the behavior of an LLM when asked to
generate new text based on a given input text and some instructions derived from
HLF statistics computed on a benchmark dataset.
To perform the experiment we must know which LLMs we are going to evaluate and
what benchmark datasets we are going to use.
We must also select the HLFs for our experiment.
Before going into details, let us outline the experiment process.

The first task is to establish a baseline.
We do so by instructing the target LLM to generate text retaining the meaning of
the input text without any further instruction.
We repeat the text generation step 10 times using the same prompt and the same
input text.
For each text generation, we compute the following statistics for each of our
chosen HLFs:

@@

A second baseline is obtained by following the same process but with a new
prompt that builds upon the first one by adding the instruction to use example
text from our benchmark dataset of choice.
The examples are three texts chosen randomly once for each benchmark dataset.
The reason behind choosing three texts is to not overrun the context window of
the LLM.
This hints at the main drawback for our approach: it uses up the context window
with system instructions.
While this isn't an issue with all models, some only allow a context window of
around 8000 tokens so we should be mindful.
We use the same example texts for all evaluated LLMs.
Same as before, we construct the baseline by conducting 10 trials using the same
prompt and input text.

The next step is to devise HLF prompts derived from both baseline prompts.
The HLF prompts contain additional instructions to generate text that would
comply with certain HLF values.
These HLF values are obtained by computing the statistics outlined above on
the entirety of the target benchmark dataset.
The HLF prompts are constructed from templates that take into account the HLF
values.
The prompt templates are available in the code repository that accompanies this
experiment~\cite{olar2024experimentcode}.
The instructions concerning all choseon HLFs are given concomitently within the
same prompt.
In a similar manner to how we generated the baselines, we run ten trials for
each case (with or without example articles from the target benchmark dataset).

We compare the first baseline with the results obtained by using the first
modified HLF prompt to understand whether we can influence LLM text generation
using HLFs.
Our reasoning is that if the LLM were influenced by the HLF prompt, the output
would show different HLF values compared to the baseline.

We now turn our attention to the second baseline.
Given that the LLM now also has example articles, it can freely immitate those
articles when generating the text.
If there is a consistent difference over the ten trials between generating text
using the baseline or using the HLF prompt, this difference can only be a result
of the LLM complying to HLF instructions.
This difference is expressed by the change of value for each HLF between the
baseline and HLF enhanced text generation.

Because of the examples present in the prompt, we perform the second experiment
with input text both from within and from outside the corpus.

@@

\begin{table}[ht]
    \setlength\tabcolsep{6pt}
    \centering
    \begin{tabular}{@{}lllll@{}}\toprule
        Name                      & Parameters  & Context Size & Elo Score & License Type        \\ \toprule
        OpenAI GPT-4o             & undisclosed & 128000       & 1287      & Proprietary         \\
        Google Gemini 1.5 Pro     & undisclosed & 1048576      & 1268      & Proprietary         \\
        Anthropic Claude 3 Opus   & undisclosed & 200000       & 1248      & Proprietary         \\
        Meta Llama 3-70B-Instruct & 70 Billion  & 8000         & 1208      & Llama 3 Community   \\
        Cohere Command R+         & 104 Billion & 128000       & 1189      & CC-BY-NC 4.0        \\
                                  &             &              &           & with Acceptable Use \\
                                  &             &              &           & Addendum            \\ \bottomrule
    \end{tabular}
    \caption{Experiment LLMs}\label{table-llm}
\end{table}

@@

Our main focus is to understand how susceptible an LLMs is to prompts telling it
to conform to HLFs.
LFTK~\cite{lftk-2023} authors categorize HLFs by domain and family.
Furthermore, information is available on the suitability of each HLF for given
tasks such as readability assessment or fake news detection.
However, there is no direct mapping between HLFs and style.
It falls onto us to choose features based on their influence over text style as
recognized in literature~\cite{verma2019lexical,lugea2023stylistics}.

Another HLF characteristic we have in mind is their level of abstraction.
Features such as the word count or number of punctuation mark are very concrete.
On the other side of the spectrum, features such as the Flesch-Kincaid grade
level~\cite{kincaid1975derivation} are highly abstract.
We divide HLFs in five categories ranging from very concrete to highly abstract
and choose 2 features from each category for a grand total of 10 HLFs.
Our final selection is available in Table~\ref{table-hlf}.

\begin{table}[ht]
    \setlength\tabcolsep{6pt}
    \centering
    \begin{tabular}{@{}lllll@{}}\toprule
        Name                              & Family           & Domain           & Abstraction     & Style Level\footnotemark \\ \toprule
        total unique words                & wordsent         & surface          & very concrete   & lexical                  \\
        total sentences                   & wordsent         & surface          & very concrete   & syntactic                \\ \bottomrule
        total number of named entities    & entity           & discourse        & concrete        & semantic                 \\
        total number of unique adjectives & partofspeech     & syntax           & concrete        & syntactic                \\ \bottomrule
        simple type token ratio           & typetokenratio   & lexico-semantics & regular         & pragmatic                \\
        number of verbs per word          & partofspeech     & surface          & regular         & syntactic                \\ \bottomrule
        corrected adjective variation     & lexicalvariation & lexico-semantics & abstract        & lexical                  \\
        corrected verb variation          & lexicalvariation & lexico-semantics & abstract        & lexical                  \\ \bottomrule
        Flesch-Kincaid grade level        & readformula      & surface          & highly abstract & pragmatic                \\
        Kuperman age of aquisition        & worddiff         & lexico-semantics & highly abstract & lexical                  \\ \bottomrule
    \end{tabular}
    \caption{Handcrafted Linguistic Features}\label{table-hlf}
\end{table}
\footnotetext{as described in the figure in~\cite{lugea2023stylistics}, chapter 1, page 7}

@@

We have run this experiment on the datasets listed in Table~\ref{table-ds}.

\begin{table}[ht]
    \setlength\tabcolsep{6pt}
    \centering
    \begin{tabular}{@{}lllll@{}}\toprule
        Name              & Task                       & Row Count & Content Type & Reference                     \\ \toprule
        Yelp Reviews Full & Sentiment Classification   & 700000    & Reviews      & \cite{yelp2015neurips}        \\
                          & Text Style Transfer        & 700000    &              &                               \\
                          & Text Classification        & 700000    &              &                               \\ \midrule
        CNN / DailyMail   & Summarization              & 300000    & News         & \cite{cnndm2015}              \\
                          & Question Answering         &           & Stories      & \cite{cnndm2017}              \\
                          & Text Generation            &           &              &                               \\ \midrule
        HANNA             & Automatic Story Generation & 1056      & Stories      & \cite{hanna2024}              \\ \midrule
        StylePTB          & Text Style Transfer        & 60000     & Sentences    & \cite{lyu-etal-2021-styleptb} \\ \bottomrule
    \end{tabular}
    \caption{Experiment LLMs}\label{table-ds}
\end{table}

The Yelp review data set provides a diverse assortment of writing styles and
thus allowing us to home in on an LLMs ability to follow HLFs instead of copying
a writing style.
The reviews are often relatively short texts written in a coloquial style with
a fair amount of variation in sentiment.

On the other hand the CNN and DailyMail stories are news stories and even though
their style might vary from one author to another, they will definitely be more
formal than Yelp reviews.
The writing is likely to be more mature, too.

The Human-ANnotated NArratives (HANNA) dataset is interesting because it
contains stories generated by LLMs. Since LLM generated stories can be assessed
just as easily as human generated stories by using HLFs, we thought it would be
a welcome addition to run our experiment for non-human writers, too.

Finally, StylePTB is a dataset containing fine grained stylistic changes at the
sentence level. We use this dataset to see how well LLMs do in scenarios
involving HLFs at the sentence level (for example, part of speech HLFs).

@@

The input prompts are derived from the templates available in the Git repository
that accompanies this article~\cite{olar2024experimentcode}.
We provide the LLM with a system prompt that contains instructions concerning
all HLFs.

@@

Handcrafted linguistic features are integer or floating point values that we
can compute on text.
All our graphs show the HLF on the vertical axis, while the horizontal axis is
reserved for the trial number.
The plots lend themselves to easy interpretation based on some simple rules.
We plot the HLF values for the text generated from baseline prompts as a
continuous grey line.
The similar values obtained by using HLF enhanced prompts are plotted using a
continuous grenat line.
To highlight the `tolerance' obtained by adding and subtracting the variance of
the HLF on the benchmark dataset, we plot light gray areas around the baseline
and HLF-enhanced plot lines.
The minimum and maximum HLF on the benchmark dataset are plotted using dotted
black lines.
A continuous black line designates the average HLF on the benchmark dataset.

Following this description of how to interpret the graphs, we present sample
findings using the features for which we've observed the best and the worst
compliance by the LLM.
These findings were all made on the <DECIDE LATER> dataset.

\subsection{Can an LLM Follow HLF Instructions?}

Yelp Results on the best and worst performing metric.
The rest of the metrics in an Appendix.

@@

Yelp results on the best and worst performing metric.
The rest of the metrics (including on all other datasets) in an Appendix.
